Plug-and-play dual-tree algorithm runtime analysis

### 2014-12-24 bounds-files/sec02-nearest-neighbor-search.tex
Refactor KDE a bit; we can write about both approximation schemes.
Ready for review.

\begin{thm}
Using cover trees, the standard cover tree pruning dual-tree traversal, and the
nearest neighbor search \texttt{BaseCase()} and \texttt{Score()} as given in
Algorithms \ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively, and also
given a reference set $S_r$ with expansion constant $c_r$, a query set $S_q$,
and an inverse degree of bichromaticity $\nu$, the running time of the
algorithm is bounded by $O(c_r^4 c_{qr}^5 \nu N)$.
\label{thm:nns}
\end{thm}

\begin{proof}
The running time of \texttt{BaseCase()} and \texttt{Score()} are clearly $O(1)$.
Due to Theorem \ref{thm:ct-runtime}, we therefore know that the runtime of the
algorithm is bounded by $O(c_r^4 \nu |R^*| N)$.  Thus, the only thing that
remains is to bound the maximum size of the reference set, $|R^*|$.

@@

In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 \nu N)$ because
$c = c_r = c_{qr}$.

>>>>

\begin{thm}
Using cover trees, the standard cover tree pruning dual-tree traversal, and the
nearest neighbor search \texttt{BaseCase()} and \texttt{Score()} as given in
Algorithms \ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively, and also
given a reference set $S_r$ with expansion constant $c_r$, a query set $S_q$,
the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
i_t(\mathscr{T}_q) + \theta))$ with $i_t(\mathscr{T}_q$ and $\theta$ defined as
in Definition \ref{def:imbalance} and Lemma \ref{lem:extcase3}, respectively.
\label{thm:nns}
\end{thm}

\begin{proof}
The running time of \texttt{BaseCase()} and \texttt{Score()} are clearly $O(1)$.
Due to Theorem \ref{thm:ct-runtime}, we therefore know that the runtime of the
algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
Thus, the only thing that remains is to bound the maximum size of the reference
set, $|R^*|$.

@@

In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
i_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
well-behaved trees where $i_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
this represents the current best worse-case runtime bound for nearest neighbor
search.

### 2014-12-24 bounds-files/sec03-approx-kde.tex
Ready for review.

\begin{equation}
| f(p_q) - f^*(p_q) | < N \epsilon.
\end{equation}

@@

\begin{equation}
\frac{| f(p_q) - f^*(p_q) |}{| f^*(p_q) |} < N \epsilon.
\end{equation}

@@

    \IF{$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) /
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) < \epsilon$}
      \STATE $f_n(\mathscr{N}_q) \gets f_n(\mathscr{N}_q) + |
\mathscr{D}^p(\mathscr{N}_r) | \left(\mathcal{K}(d_{\min}(\mathscr{N}_q,
\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right) /\;2$
      \RETURN $\infty$
    \ENDIF

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
of the previous section.  Then, given a query set $S_q$ and a reference set
$S_r$ with expansion constant $c_r$, and using the approximate kernel density
estimation \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
\ref{alg:kde_base_case} and \ref{alg:kde_score}, respectively, with the
traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
approximate kernel density estimation for some $\epsilon$ error parameter is
bounded by
%\begin{equation}
$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + \theta))$
%\end{equation}
where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$ and
$\theta$ as in Theorem \ref{thm:ct-runtime}.

@@

\begin{proof}
It is clear that \texttt{BaseCase()} and \texttt{Score()} both take $O(1)$ time,
so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
algorithm is bounded by $O(c_r^4 |R^*| (N + \theta))$.  Thus, we will bound
$|R^*|$ using techniques related to those used by \citet{ram2009}.  The bounding
of $|R^*|$ is split into two sections: first, show that when the scale
$s_r^{\max}$ is small enough, $R^*$ is empty.  Second, bound $R^*$ when
$s_r^{\max}$ is larger.

@@

Approximate kernel density estimation using relative-value approximation may be
bounded in a similar manner.  A \texttt{Score()} function is given in Algorithm
\ref{alg:kde_rva_score}.

>>>>

\begin{equation}
\label{eqn:ava}
| f(p_q) - f^*(p_q) | < N \epsilon.
\end{equation}

@@

\begin{equation}
\label{eqn:rva}
\frac{| f(p_q) - f^*(p_q) |}{| f^*(p_q) |} < N \epsilon.
\end{equation}

@@

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
of the previous section.  Then, given a query set $S_q$ and a reference set
$S_r$ with expansion constant $c_r$, and using the approximate kernel density
estimation \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
\ref{alg:kde_base_case} and \ref{alg:kde_score}, respectively, with the
traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
approximate kernel density estimation for some $\epsilon$ error parameter is
bounded by
%\begin{equation}
$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + i_t(\mathscr{T}_q) + \theta))$
%\end{equation}
where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$,
$i_t(\mathscr{T}_q)$ defined as in Definition \ref{def:imbalance}, and $\theta$
defined as in Lemma \ref{lem:extcase3}.

@@

\begin{proof}
It is clear that \texttt{BaseCase()} and \texttt{Score()} both take $O(1)$ time,
so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
Thus, we will bound $|R^*|$ using techniques related to those used by
\citet{ram2009}.  The bounding of $|R^*|$ is split into two sections: first,
show that when the scale $s_r^{\max}$ is small enough, $R^*$ is empty.  Second,
bound $R^*$ when $s_r^{\max}$ is larger.

@@

Approximate kernel density estimation using relative-value approximation may be
bounded by reducing the absolute-value approximation algorithm (in linear time
or less) to relative-value approximation.  This is the same strategy as
performed by \citet{ram2009}.

First, we must establish a \texttt{Score()} function for relative value
approximation.  The difference between Equations \ref{eqn:ava} and \ref{eqn:rva}
is the division by the term $|f^*(p_q)|$.  But we can quickly bound
$|f^*(p_q)|$:

\begin{equation}
f^*(p_q) \ge N \mathcal{K}\left(\max_{p_r \in S_r} d(p_q, p_r)\right).
\end{equation}

This is clearly true, because each point in $S_r$ must contribute more than
$\mathcal{K}(\max_{p_r \in S_r} d(p_q, p_r))$ to $f^*(p_q)$.  Now, we may revise
the relative approximation condition in Equation \ref{eqn:rva}:

\begin{equation}
| f(p_q) - f^*(p_q) | \le \epsilon \mathcal{K}^{\max}
\end{equation}

\noindent where $\mathcal{K}^{\max}$ is lower bounded by $\mathcal{K}(\max_{p_r
\in S_r} d(p_q, p_r))$.  This allows us to create a \texttt{Score()} algorithm,
given in Algorithm \ref{alg:kde_rva_score}.

    \IF{$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) < \epsilon
\mathcal{K}^{\max}$}
\label{alg:rva-kde-prune}
      \STATE $f_n(\mathscr{N}_q) \gets f_n(\mathscr{N}_q) + |
\mathscr{D}^p(\mathscr{N}_r) | \left(\mathcal{K}(d_{\min}(\mathscr{N}_q,
\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right)
/\;2$ \label{alg:rva-kde-update}
      \RETURN $\infty$
    \ENDIF

Using this, we may prove linear runtime bounds for relative value approximate
kernel density estimation.

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
of the previous section.  Then, given a query set $S_q$ and a reference set
$S_r$ both of size $O(N)$, it is possible to perform relative value
approximate kernel density estimation (satisfying the condition of Equation
\ref{eqn:rva}) in $O(N)$ time, assuming that the expansion constant $c_r$ of
$S_r$ is not dependent on $N$.
\end{thm}

\begin{proof}
It is easy to see that Theorem \ref{thm:kde-bound} may be adapted to the very
slightly different \texttt{Score()} rule of Algorithm \ref{alg:kde_rva_score}
while still providing an $O(N)$ bound.  With that \texttt{Score()} function, the
dual-tree algorithm will return relative-value approximate kernel density
estimates satisfying Equation \ref{eqn:rva}.

We now turn to the calculation of $\mathcal{K}^{\max}$.  Given the cover trees
$\mathscr{T}_q$ and $\mathscr{T}_r$ with root nodes $\mathscr{N}_{r}^{R}$ and
$\mathscr{N}_{r}^{R}$, respectively, we may calculate a suitable
$\mathcal{K}^{\max}$ value in constant time:

\begin{equation}
\mathcal{K}^{\max} = d_{\max}(\mathscr{N}_q^R, \mathscr{N}_r^R) = d(p_q^R,
p_r^R) + 2^{s_q^{\max} + 1} + 2^{s_r^{\max} + 1}.
\end{equation}

This proves the statement of the theorem.
\end{proof}

In this case, we have not shown tighter bounds because the algorithm we have
proposed is not useful in practice.  For an example of a better relative-value
approximate kernel density estimation dual-tree algorithm, see the work of
\citet{gray2003nonparametric}.

### 2014-12-24 bounds-files/sec04-range-search.tex
Propagate changed Lemma 1 to the document.
Ready for review.

\begin{equation}
O\left(c_r^{4} \max\left(c_r^{4 + \beta}, |S_{\max}| + C\right) (N + \theta)
\right)
\end{equation}

with $\theta$ defined as in Theorem
\ref{thm:ct-runtime},
$\beta = \log_2 (1 + \alpha^{-1})$, and $S_{\max}$ as defined in Equation
\ref{eqn:smax}.
%\end{equation}
\label{thm:rs}
\end{thm}

\begin{proof}
Both \texttt{BaseCase()} (Algorithm \ref{alg:rs_bc}) and \texttt{Score()}
(Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
\ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
$O(c_r^4 |R^*| (N + \theta))$.  As with the previous proofs, then, our only task
is to bound the maximum size of the reference set, $|R^*|$.

@@

\begin{equation}
|R^*| \le c_r^{4 + \log_2(1 + \alpha^{-1})}.
\end{equation}

@@

\begin{equation}
O(c_r^{8 + \beta} (N + \theta)).
\end{equation}
\label{cor:rs}
\end{cor}

@@

In this setting we can more easily consider the relation of the running time to
$\alpha$.  Consider $\alpha = (1 / 3)$; this yields a running time of $O(c^8 (N
+ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + \theta))$, $\alpha = (1 /
15)$ yields $O(c^{10} (N + \theta))$, and so forth.  As $\alpha$ gets smaller,
the exponent on $c$ gets larger, and diverges as $\alpha \to 0$.

>>>>

\begin{equation}
O\left(c_r^{4} \max\left(c_r^{4 + \beta}, |S_{\max}| + C\right) (N +
i_t(\mathscr{N}_q) + \theta)
\right)
\end{equation}

with $\theta$ defined as in Lemma \ref{lem:extcase3},
$\beta = \lceil \log_2 (1 + \alpha^{-1}) \rceil$, and $S_{\max}$ as defined in
Equation \ref{eqn:smax}.
%\end{equation}
\label{thm:rs}
\end{thm}

\begin{proof}
Both \texttt{BaseCase()} (Algorithm \ref{alg:rs_bc}) and \texttt{Score()}
(Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
\ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
$O(c_r^4 |R^*| (N + i_t(\mathscr{N}_q) + \theta))$.  As with the previous
proofs, then, our only task is to bound the maximum size of the reference set,
$|R^*|$.

@@

\begin{equation}
|R^*| \le c_r^{4 + \lceil \log_2(1 + \alpha^{-1}) \rceil}.
\end{equation}

@@

\begin{equation}
O(c_r^{8 + \beta} (N + i_t(\mathscr{N}_q) + \theta)).
\end{equation}
\label{cor:rs}
\end{cor}

@@

In this setting we can more easily consider the relation of the running time to
$\alpha$.  Consider $\alpha = (1 / 3)$; this yields a running time of $O(c^8 (N
+ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + i_t(\mathscr{N}_q +
\theta))$, $\alpha = (1 / 15)$ yields $O(c^{10} (N + i_t(\mathscr{N}_q) +
\theta))$, and so forth.  As $\alpha$ gets smaller, the exponent on $c$ gets
larger, and diverges as $\alpha \to 0$.
